Full Brain Blood-Oxygen-Level-Dependent Signal Parameter Estimation Using Particle Filters

### 2010-08-20 doc/paper/main.tex
More in the introduction

\section{Introduction}
\label{sec:Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{raditional} methods of analyzing 
Functional Magnetic Resonance Imaging (FMRI)
time series perform regression using a linear 
combination of static explanatory variables. 
In this paper I will demonstrate the use of the Particle 
Filters as a means of estimating the governing parameters of the 
nonlinear BOLD model at a computation cost 
that would still allow real time calculations for multiple voxels.
Practically, this paper presents an alternative
to detecting neural activity from the traditionally
used Statistical Parametric Mapping (SPM). Though more computationally intense,
this method is capable of modeling nonlinear effects, 
and provides significantly more detailed output. 
By using a separate particle filter for each single time series it 
is possible to estimate parameters and make real-time predictions
for small neural regions, a feature which could be useful towards real time FMRI 
\cite{DeCharms2005}. Future works will also benefit from the ability to 
apply conditions to the posterior distribution in post-processing without
recalculating parameters; for instance to impose additional 
physiological
constraints. Modeling the BOLD (Blood-Oxygen-Level-Dependent) response 
as a nonlinear system is the
best way to determine the correlation of stimulus sequence with the BOLD
response; yet in the past doing this on a large scale has been far too
computationally taxing. The solution used here takes approximately 40 seconds
for a single 5 minute time series (with Core 2 Duo Q6600).

>>>>

\section{Introduction}
\label{sec:Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{raditional} methods of analyzing 
Functional Magnetic Resonance Imaging (FMRI)
time series perform regression using a linear 
combination of static explanatory variables. 
However, the static nature of the General Linear Model (GLM)
limits its potential use. It is well known that different 
Hemodynamic Response Functions are necessary 
for different regions of the brain to prevent excessive false
negatives \cite{Handwerker2004}. Additionally quite a few studies
have reported spatially varying nonlinearities \cite{Wager2005,Birn2001}.
Besides not allowing HRF differences between patients, there is no
reasonable way to incorporate other forms of physiological
data. Combined FMRI CBF or CBV imaging methods are improving,
as seen in Chen et al. \cite{Chen2009}. These techniques could shed light on
neural activation by providing extra measurements, yet a 
physiologically reasonable model is necessary to incorporate this extra data.
Activation detection methods also don't have the ability 
to identify pathologies based on state variables or parameters. For
example, decreased compliance of blood vessels could indicate, or 
even cause, a neurological condition that 
is not easily seen in other imaging modalities.

It is well known that the changes in deoxy-hemoglobin content is the
primary driver of short term changes in MR signal for FMRI imaging
techniques \cite{Buxton1998, WEISSKOFF1994, Ogawa, Obata2004}. For
this reason, the nonlinear state equations that change Deoxy-Hb (DHb) 
content have been heavily studied, and are well-characterized. Attempts
at learning the parameters of the BOLD model have also been actively
studied but significantly less successful. The most wide-spread method
of calculating the parameters, from two papers by Friston et al., 
are part of the SPM toolkit; though this algorithm is rarely used 
as well \cite{Friston2000, Friston2002b}. 
This method depends on a quadratic convolution (two term Volterra Kernel)
based estimation of the BOLD output to obtain partials with respect 
to parameters. Problematically, quality of the Volterra estimate
is not well known, so it is difficult to quantify error. This method 
also doesn't account for noise in the parameters or the state variables.
These shortcomings have led to several other separate attempts. In particular 
Riera et al. linearized the noise and thus performed regression between
the model and steps of the BOLD output \cite{Riera2003}. 
In Vakorin et al., a combination of Genetic Algorithms and Simulated 
Annealing were used to estimate not only the parameters, but the 
true stimuli driving the BOLD signal \cite{Vakorin2007}. 
This addresses the inherent uncertainty of exactly where and when 
stimuli actually get applied. Unfortunately this algorithm was extremely slow.
In Johnston et al. a particle filter was applied with the purpose of
estimating the output distribution, which was then used to optimize
parameters by means of least squares based techniques \cite{Johnston2008}. 
Murray used a particle filter, similar to \cite{Johnston2008}, but
held the model parameters constant; thus all error in the output
was treated as the result of noise in the underlying parameters 
\cite{Murray2008}.  Hu et al. used an unscented kalman filter (UKF) 
was used to optimize parameters \cite{Hu2009}.

\section{Methods}
\label{sec:Methods}
\begin{itemize}
\item BOLD Model
\item Particle Filter Description 
\item Description of Simulation
\item Description of Real FMRI data
\end{itemize}

\section{Results}
\label{sec:Results}
\begin{itemize}
\item Simulation Results - Possum Only
\begin{itemize}
    \item Histogram of parameters for each region
    \item Final histogram of one run?
    \item Comparison of MI with SNR
\end{itemize}
\item FMRI Results 
\begin{itemize}
    \item Activation maps comparing M.I. with SPM
\end{itemize}
\end{itemize}

\section{Discussion}
\label{sec:Discussion}

\begin{itemize}
\item Parameters ill-defined
\end{itemize}

\section{Conclusion}
\label{sec:Conclusion}
